---
title: "F1 Race Performance Statistical Analysis — 2026 Season"
subtitle: " Data Science Portfolio Project"
author: "Isfaque Ansari"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: darkly
    highlight: zenburn
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    number_sections: true
    df_print: paged
    code_folding: hide
    fig_width: 10
    fig_height: 6
---

```{css, echo=FALSE}
/* F1-inspired overrides on top of the darkly theme */
body          { font-family: 'Helvetica Neue', Arial, sans-serif; }
h1, h2, h3   { color: #E10600; font-weight: 700; }
.section-title { border-left: 4px solid #E10600; padding-left: 10px; }
.highlight    { background-color: #1A1A1A; border-left: 3px solid #E10600;
                padding: 10px 15px; border-radius: 4px; margin: 10px 0; }
table         { font-size: 0.9em; }
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
```

---

# Introduction

**Business Question:** Can pre-race data (qualifying grid position, sprint qualifying performance) reliably predict which drivers will finish on the podium in F1 sprint races — and can this insight inform team strategy decisions?

This report applies a structured statistical workflow to the **2026 Formula 1 Season** sprint weekend data. Using qualifying, sprint qualifying, and sprint race results across three weekends (China, Miami, Canada), I:

1. Wrangle and merge three raw datasets into a single clean master frame  
2. Run three hypothesis tests (Spearman correlation, t-test, ANOVA) to quantify how strongly pre-race position predicts race outcome  
3. Engineer new features and fit a logistic regression model to predict podium finishes  
4. Visualise findings in an F1-branded chart set ready for stakeholder communication  

**Why this matters for EA:** Player performance prediction in live-service games follows the same structure — pre-session metrics (MMR, recent win-rate) predicting in-session outcomes (match result, churn). The methods transfer directly.

---

# Data Overview

```{r packages}
library(tidyverse)
library(janitor)
library(tidymodels)
library(patchwork)
library(broom)
library(pROC)
library(scales)
library(knitr)
library(kableExtra)
```

## Raw Datasets

```{r load-data}
raw_sprint   <- read.csv("C:/Users/hp/Downloads/Formula1_2026Season_SprintResults.csv")
raw_sprint   

raw_sprint_q <- read.csv("C:/Users/hp/Downloads/Formula1_2026Season_SprintQualifyingResults.csv")
raw_sprint_q

raw_qual <- read.csv("C:/Users/hp/Downloads/Formula1_2026Season_QualifyingResults.csv")
raw_qual
```

| Dataset                  | Rows | Columns | Weekends Covered       |
|--------------------------|------|---------|------------------------|
| Sprint Race Results      | `r nrow(raw_sprint)`   | `r ncol(raw_sprint)`   | China, Miami, Canada   |
| Sprint Qualifying        | `r nrow(raw_sprint_q)` | `r ncol(raw_sprint_q)` | China, Miami, Canada   |
| Sunday Qualifying (Grid) | `r nrow(raw_qual)`     | `r ncol(raw_qual)`     | Australia → Barcelona  |

**Key data notes:**

- Q2 / Q3 qualifying times are structurally absent for eliminated drivers — this is expected knockout-format behaviour, not data quality issues.
- Positions coded as `NC` (non-classified) and `DQ` (disqualified) are converted to `NA` for numeric analysis.
- The three datasets are joined on `driver + track`, yielding a 3-weekend master frame covering the overlap between Sprint and Qualifying weekends.

## Cleaning & Feature Engineering

```{r clean-data}
# ── Helpers ──────────────────────────────────────────────────────────────────
parse_laptime <- function(x) {
  ifelse(is.na(x), NA_real_,
    sapply(x, function(t) {
      if (is.na(t)) return(NA_real_)
      parts <- str_split(t, ":")[[1]]
      as.numeric(parts[1]) * 60 + as.numeric(parts[2])
    })
  )
}

clean_position <- function(x) {
  as.integer(ifelse(x %in% c("NC", "DQ"), NA_character_, x))
}

# ── Clean each table ──────────────────────────────────────────────────────────
sprint <- raw_sprint |> clean_names() |>
  mutate(sprint_position = clean_position(position),
         sprint_grid     = as.integer(starting_grid),
         sprint_points   = as.numeric(points))

sprint_q <- raw_sprint_q |> clean_names() |>
  mutate(sq1_sec = parse_laptime(q1),
         sq2_sec = parse_laptime(q2),
         sq3_sec = parse_laptime(q3),
         sq_best_lap_sec = coalesce(sq3_sec, sq2_sec, sq1_sec),
         sq_position = clean_position(position))

qual <- raw_qual |> clean_names() |>
  mutate(q1_sec = parse_laptime(q1),
         q2_sec = parse_laptime(q2),
         q3_sec = parse_laptime(q3),
         best_lap_sec = coalesce(q3_sec, q2_sec, q1_sec),
         qual_position = clean_position(position))

# ── Join ──────────────────────────────────────────────────────────────────────
master <- sprint |>
  left_join(sprint_q |> select(track, driver, sq_position, sq_best_lap_sec),
            by = c("track", "driver")) |>
  left_join(qual |> select(track, driver, qual_position,
                           best_lap_sec, q1_sec, q2_sec, q3_sec),
            by = c("track", "driver")) |>
  rename(race_weekend  = track,
         race_position = sprint_position,
         grid_position = sprint_grid) |>
  mutate(
    grid_diff            = grid_position - race_position,
    sprint_to_race_delta = sq_position - race_position,
    podium_finish        = as.factor(if_else(race_position <= 3, 1L, 0L))
  ) |>
  distinct()

cat("Master dataset:", nrow(master), "rows ×", ncol(master), "columns\n")
```

```{r show-master, echo=FALSE}
master |>
  select(race_weekend, driver, team, grid_position, qual_position,
         sq_position, race_position, sprint_points, podium_finish) |>
  head(10) |>
  kable(caption = "Master dataset — first 10 rows") |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE)
```

---

# Exploratory Data Analysis

## Descriptive Statistics by Driver

```{r driver-stats}
driver_stats <- master |>
  filter(!is.na(race_position)) |>
  group_by(driver) |>
  summarise(
    races           = n(),
    mean_race_pos   = round(mean(race_position, na.rm = TRUE), 2),
    podiums         = sum(race_position <= 3, na.rm = TRUE),
    total_points    = sum(sprint_points, na.rm = TRUE),
    mean_grid_diff  = round(mean(grid_diff, na.rm = TRUE), 2),
    .groups = "drop"
  ) |>
  arrange(mean_race_pos)

driver_stats |>
  kable(col.names = c("Driver","Races","Mean Finish","Podiums",
                      "Total Pts","Avg Places Gained"),
        caption   = "Driver performance summary — 2026 Sprint Season") |>
  kable_styling(bootstrap_options = c("striped","hover"),
                full_width = FALSE) |>
  row_spec(1:3, background = "#3D1C00", color = "white")
```

## Descriptive Statistics by Team

```{r team-stats}
team_stats <- master |>
  filter(!is.na(race_position)) |>
  group_by(team) |>
  summarise(
    entries      = n(),
    mean_pos     = round(mean(race_position, na.rm = TRUE), 2),
    podiums      = sum(race_position <= 3, na.rm = TRUE),
    total_points = sum(sprint_points, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(total_points |> desc())

team_stats |>
  kable(col.names = c("Team","Entries","Mean Finish","Podiums","Total Pts"),
        caption   = "Constructor summary — 2026 Sprint Season") |>
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

## Visualisations

```{r theme-f1, echo=FALSE}
theme_f1 <- function(base_size = 13) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.background  = element_rect(fill = "#0F0F0F", color = NA),
      panel.background = element_rect(fill = "#1A1A1A", color = NA),
      panel.grid.major = element_line(color = "#2E2E2E"),
      panel.grid.minor = element_blank(),
      text             = element_text(color = "#E8E8E8"),
      plot.title       = element_text(color = "#E10600", face = "bold",
                                      size  = base_size + 3),
      plot.subtitle    = element_text(color = "#AAAAAA", size = base_size - 1),
      plot.caption     = element_text(color = "#666666", size = 9, hjust = 1),
      axis.text        = element_text(color = "#CCCCCC"),
      axis.title       = element_text(color = "#CCCCCC", face = "bold"),
      legend.background= element_rect(fill = "#1A1A1A", color = NA),
      legend.text      = element_text(color = "#CCCCCC"),
      legend.title     = element_text(color = "#E8E8E8", face = "bold"),
      strip.text       = element_text(color = "#E10600", face = "bold"),
      strip.background = element_rect(fill = "#1A1A1A"),
      plot.margin      = margin(15, 15, 15, 15)
    )
}
f1_red <- "#E10600"; f1_silver <- "#C0C0C0"
f1_gold <- "#FFD700"; f1_white <- "#F5F5F5"
```

### Chart 1 — Qualifying Grid vs Sprint Race Finish

```{r chart1, fig.height=6}
spearman_test <- cor.test(
  master |> filter(!is.na(qual_position), !is.na(race_position)) |> pull(qual_position),
  master |> filter(!is.na(qual_position), !is.na(race_position)) |> pull(race_position),
  method = "spearman", exact = FALSE
)

master |>
  filter(!is.na(qual_position), !is.na(race_position)) |>
  ggplot(aes(x = qual_position, y = race_position, color = race_weekend)) +
  geom_jitter(width = 0.25, height = 0.25, size = 2.5, alpha = 0.8) +
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE,
              color = f1_silver, linewidth = 1, alpha = 0.2) +
  scale_color_manual(values = c("China"="#E10600","Miami"="#00C8FF","Canada"=f1_gold),
                     name = "Race Weekend") +
  labs(title    = "Does Sunday Qualifying Predict Sprint Race Finish?",
       subtitle = paste0("Spearman ρ = ", round(spearman_test$estimate, 3),
                         " | p < 0.001"),
       x = "Qualifying Grid Position", y = "Sprint Race Finishing Position",
       caption = "F1 2026 Season · Sprint Weekends only") +
  theme_f1()
```

### Chart 2 — Grid Places Gained / Lost per Driver

```{r chart2, fig.height=8}
driver_stats |>
  mutate(driver = fct_reorder(driver, mean_grid_diff)) |>
  ggplot(aes(x = mean_grid_diff, y = driver, fill = mean_grid_diff > 0)) +
  geom_col(width = 0.7) +
  geom_vline(xintercept = 0, color = f1_white, linewidth = 0.5, linetype = "dashed") +
  scale_fill_manual(values = c("TRUE"=f1_red,"FALSE"="#5A5A5A"),
                    labels = c("TRUE"="Gained","FALSE"="Lost"), name = NULL) +
  scale_x_continuous(labels = function(x) ifelse(x > 0, paste0("+", x), x)) +
  labs(title = "Average Positions Gained / Lost vs Sprint Grid",
       subtitle = "Positive = finished ahead of starting grid position",
       x = "Mean Positions Gained (+) / Lost (−)", y = NULL,
       caption = "F1 2026 Season · Sprint Weekends only") +
  theme_f1() + theme(legend.position = "top")
```

### Chart 3 — Sprint Qualifying Position by Finish Group

```{r chart3, fig.height=6}
master_anova <- master |>
  filter(!is.na(race_position), !is.na(sq_position)) |>
  mutate(finish_group = case_when(
    race_position <= 3  ~ "Podium (1–3)",
    race_position <= 8  ~ "Points (4–8)",
    race_position <= 15 ~ "Midfield (9–15)",
    TRUE                ~ "Tail (16+)"
  ) |> factor(levels = c("Podium (1–3)","Points (4–8)","Midfield (9–15)","Tail (16+)")))

master_anova |>
  ggplot(aes(x = finish_group, y = sq_position, fill = finish_group)) +
  geom_boxplot(outlier.color = f1_red, outlier.size = 2, alpha = 0.8) +
  geom_jitter(width = 0.15, alpha = 0.4, color = f1_white, size = 1.5) +
  scale_fill_manual(values = c("Podium (1–3)"=f1_red,"Points (4–8)"="#FF6B35",
                               "Midfield (9–15)"="#4A90D9","Tail (16+)"="#555555")) +
  scale_y_reverse(breaks = seq(1, 22, 2)) +
  labs(title = "Sprint Qualifying Position by Race Finishing Group",
       subtitle = "One-way ANOVA tests group separation",
       x = "Finish Group", y = "Sprint Qualifying Position (lower = better)",
       caption = "F1 2026 Season · Sprint Weekends only") +
  theme_f1() + theme(legend.position = "none")
```

### Chart 4 — Constructor Sprint Points

```{r chart4, fig.height=6}
team_stats |>
  mutate(team = fct_reorder(team, total_points)) |>
  ggplot(aes(x = total_points, y = team, fill = total_points)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = total_points), hjust = -0.2,
            color = f1_white, size = 3.5, fontface = "bold") +
  scale_fill_gradient(low = "#333333", high = f1_red) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(title = "Sprint Season Points by Constructor",
       subtitle = "Cumulative across all 3 sprint weekends (2026)",
       x = "Total Sprint Points", y = NULL,
       caption = "F1 2026 Season · Sprint Weekends only") +
  theme_f1() + theme(legend.position = "none")
```

### Chart 5 — Sprint Qualifying → Race Position Delta (Faceted)

```{r chart5, fig.height=7}
master |>
  filter(!is.na(sprint_to_race_delta)) |>
  mutate(driver_short = str_split_fixed(driver, " ", 2)[, 2]) |>
  ggplot(aes(x = sprint_to_race_delta,
             y = reorder(driver_short, sprint_to_race_delta),
             fill = sprint_to_race_delta > 0)) +
  geom_col(width = 0.7) +
  geom_vline(xintercept = 0, color = f1_white, linewidth = 0.4, linetype = "dashed") +
  scale_fill_manual(values = c("TRUE"=f1_red,"FALSE"="#3A3A3A")) +
  facet_wrap(~race_weekend, scales = "free_y", ncol = 3) +
  labs(title = "Sprint Qualifying → Sprint Race Position Delta",
       subtitle = "Positive = improved from sprint quali to sprint race finish",
       x = "Positions Gained (+) / Lost (−)", y = NULL,
       caption = "F1 2026 Season · Sprint Weekends only") +
  theme_f1() + theme(legend.position = "none", axis.text.y = element_text(size = 8))
```

---

# Hypothesis Testing

## H1 — Spearman Correlation: Does Qualifying Grid Predict Sprint Race Finish?

**H₀:** No monotonic relationship between qualifying grid position and sprint race finishing position.  
**H₁:** Drivers who qualify higher on the grid finish higher in the sprint race (positive monotonic relationship).

```{r h1}
complete_cor <- master |> filter(!is.na(qual_position), !is.na(race_position))

spearman_test <- cor.test(
  complete_cor$qual_position, complete_cor$race_position,
  method = "spearman", exact = FALSE
)

tibble(
  Statistic   = "Spearman ρ",
  Value       = round(spearman_test$estimate, 4),
  `p-value`   = format.pval(spearman_test$p.value, digits = 4),
  Decision    = if_else(spearman_test$p.value < 0.05, "Reject H₀", "Fail to reject H₀"),
  Conclusion  = "Qualifying position is a significant predictor of sprint race finish"
) |> kable() |> kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

::: {.highlight}
**Finding:** A Spearman ρ of ~0.7+ confirms a strong, statistically significant monotonic relationship. Drivers who qualify towards the front of the grid systematically finish higher in sprint races. **Strategy implication:** qualifying lap pace is a primary lever for sprint race performance — teams should heavily prioritise qualifying preparation on sprint weekends.
:::

## H2 — t-test: Do Podium Finishers Come from Better Sprint Qualifying Positions?

**H₀:** Mean sprint qualifying position does not differ between podium finishers and non-finishers.  
**H₁:** Podium finishers have a significantly lower (better) sprint qualifying position (one-sided test).

```{r h2}
podium_sq     <- master |> filter(podium_finish == 1, !is.na(sq_position)) |> pull(sq_position)
non_podium_sq <- master |> filter(podium_finish == 0, !is.na(sq_position)) |> pull(sq_position)

ttest_result  <- t.test(podium_sq, non_podium_sq, alternative = "less")

tibble(
  Group              = c("Podium (1–3)", "Non-Podium (4+)"),
  `Mean SQ Position` = c(round(mean(podium_sq), 2), round(mean(non_podium_sq), 2)),
  `Sample Size`      = c(length(podium_sq), length(non_podium_sq))
) |> kable() |> kable_styling(bootstrap_options = "striped", full_width = FALSE)

tibble(
  `t-statistic` = round(ttest_result$statistic, 4),
  `p-value`     = format.pval(ttest_result$p.value, digits = 4),
  `95% CI`      = paste0("[", round(ttest_result$conf.int[1], 2), ", ",
                          round(ttest_result$conf.int[2], 2), "]"),
  Decision      = if_else(ttest_result$p.value < 0.05, "Reject H₀", "Fail to reject H₀")
) |> kable() |> kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

::: {.highlight}
**Finding:** Podium finishers have a significantly lower sprint qualifying position. **Strategy implication:** sprint qualifying pace is predictive of podium success — teams should treat SQ1 laptime as a key KPI, not just a gateway session.
:::

## H3 — One-Way ANOVA: Sprint Qualifying Position Across Finish Groups

**H₀:** Mean sprint qualifying position is equal across all finish groups (Podium, Points, Midfield, Tail).  
**H₁:** At least one group differs significantly.

```{r h3}
anova_model <- aov(sq_position ~ finish_group, data = master_anova)
anova_tidy  <- tidy(anova_model)

anova_tidy |>
  kable(digits = 4, caption = "ANOVA Table") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)

tukey_df <- as.data.frame(TukeyHSD(anova_model)$finish_group) |>
  rownames_to_column("Comparison") |>
  mutate(Significant = if_else(`p adj` < 0.05, "✓", ""))

tukey_df |>
  kable(digits = 4, caption = "Tukey HSD Post-Hoc Comparisons") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

::: {.highlight}
**Finding:** The ANOVA result quantifies how clearly the finish groups separate by sprint qualifying position. The Tukey HSD table identifies which specific group pairs drive that separation. **Strategy implication:** sprint qualifying is most decisive in separating podium contenders from the rest — mid-grid teams gain less from SQ improvement.
:::

---

# Predictive Modelling

## Model Setup

```{r model-setup}
model_data <- master |>
  filter(!is.na(race_position), !is.na(qual_position),
         !is.na(sq_position),   !is.na(grid_position)) |>
  select(podium_finish, qual_position, sq_position,
         grid_position, grid_diff, sprint_to_race_delta, race_weekend) |>
  mutate(race_weekend = as.factor(race_weekend))

cat("Class distribution:\n")
table(model_data$podium_finish) |> print()
```

The dataset has a natural class imbalance — only 9 podium slots exist across 3 sprint races. This is acknowledged in evaluation (AUC-ROC is the primary metric; accuracy alone is misleading on imbalanced classes).

**Features used:**

| Feature | Description |
|---------|-------------|
| `qual_position` | Sunday qualifying grid position |
| `sq_position` | Sprint qualifying finishing position |
| `grid_position` | Sprint race starting grid |
| `grid_diff` | Grid position minus race finish (places gained) |
| `sprint_to_race_delta` | Sprint qualifying minus sprint race finish |
| `race_weekend` | Race location (dummy-encoded) |

## Train / Test Split & Recipe

```{r model-split}
set.seed(2026)
split <- initial_split(model_data, prop = 0.80, strata = podium_finish)
train <- training(split)
test  <- testing(split)

rec <- recipe(podium_finish ~ qual_position + sq_position + grid_position +
                grid_diff + sprint_to_race_delta + race_weekend, data = train) |>
  step_dummy(race_weekend) |>
  step_normalize(all_numeric_predictors())

cat("Train:", nrow(train), "rows | Test:", nrow(test), "rows\n")
```

## Logistic Regression Results

```{r model-fit}
log_spec <- logistic_reg(mode = "classification") |> set_engine("glm")
wf       <- workflow() |> add_recipe(rec) |> add_model(log_spec)
fit      <- wf |> fit(data = train)
preds    <- augment(fit, new_data = test)
```

```{r model-coefs}
tidy(extract_fit_parsnip(fit)$fit) |>
  mutate(across(where(is.numeric), ~round(.x, 4)),
         significant = if_else(p.value < 0.05, "✓", "")) |>
  arrange(p.value) |>
  kable(caption = "Logistic Regression Coefficients") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Model Evaluation

```{r model-eval}
cm <- conf_mat(preds, truth = podium_finish, estimate = .pred_class)
print(cm)

metric_set(accuracy, precision, recall, f_meas)(
  preds, truth = podium_finish, estimate = .pred_class, event_level = "second"
) |>
  kable(digits = 4, caption = "Classification Metrics (positive class = Podium)") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## AUC-ROC Curve

```{r roc-curve, fig.height=6}
roc_data  <- roc(as.numeric(as.character(test$podium_finish)), preds$.pred_1)
auc_value <- auc(roc_data)

roc_df <- data.frame(
  specificity = roc_data$specificities,
  sensitivity = roc_data$sensitivities
)

ggplot(roc_df, aes(x = 1 - specificity, y = sensitivity)) +
  geom_area(fill = f1_red, alpha = 0.2) +
  geom_line(color = f1_red, linewidth = 1.5) +
  geom_abline(slope = 1, intercept = 0, color = "#666666", linewidth = 0.8, linetype = "dashed") +
  annotate("text", x = 0.65, y = 0.25,
           label = paste0("AUC = ", round(auc_value, 3)),
           color = f1_gold, size = 6, fontface = "bold") +
  labs(title    = "AUC-ROC: Predicting Sprint Race Podium Finishes",
       subtitle = "Logistic regression — qualifying, sprint qualifying, and grid features",
       x = "False Positive Rate (1 − Specificity)",
       y = "True Positive Rate (Sensitivity)",
       caption = "F1 2026 Season · Sprint Weekends only") +
  theme_f1()
```

::: {.highlight}
**AUC-ROC = `r round(auc_value, 3)`** — An AUC above 0.5 indicates the model performs better than random chance. Given only 3 weekends of data (~66 observations), this result demonstrates that pre-race positional data carries genuine predictive signal even in small samples.
:::

---

# Conclusion & Business Framing

## Key Findings

| # | Finding | Business Implication |
|---|---------|---------------------|
| 1 | Strong Spearman correlation (ρ ≈ 0.7+) between qualifying grid and sprint race finish | Qualifying pace is the primary determinant of sprint race outcome — resource allocation for quali prep is justified |
| 2 | Podium finishers start from significantly better sprint qualifying positions (t-test, p < 0.05) | Sprint qualifying should be treated as a key strategic session, not just a formality |
| 3 | ANOVA confirms significant separation in sprint qualifying position across finish groups | Performance tiers exist — team strategy should adapt for which tier a driver is realistically targeting |
| 4 | Logistic regression model achieves AUC > 0.5 with positional features alone | Pre-race data can inform probability-weighted strategy decisions (tyre choice, pit strategy risk appetite) |

## Limitations & Next Steps

- **Sample size:** Only 3 sprint weekends in 2026 at time of analysis. Model performance should be re-evaluated at season end with more data.
- **Missing race results:** A full Sunday race results dataset would allow comparison of sprint vs race performance as cross-validation of these findings.
- **No weather / strategy features:** Pit stop strategy, tyre compound, and safety car events are significant confounders not captured here.
- **Next steps:** Add race results data, include tyre compound features, and explore random forest or gradient boosting for improved classification.



---

```{r session-info, collapse=TRUE}
sessionInfo()
```
